---
title: "Weathering Storms - which types of events are most harmful to population health or have the greatest economic consequences?"
author: "Kirsten Meeker"
date: "September 28, 2016"
output: html_document
---

```{r setup, include=FALSE, echo=TRUE}
knitr::opts_chunk$set(echo = TRUE)
```

## Synposis
This data analysis attempts to address the following questions:   
1. Across the United States, which types of events are most harmful with respect to population health?   
2. Across the United States, which types of events have the greatest economic consequences?   

## Finding the data to answer the questions
Download file from course link: https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2

```{r load_data, cache = TRUE}
df = read.csv(bzfile("repdata_data_StormData.csv.bz2"))
```
The assignment asks for the event types across the U.S., and gives EVTYPE as the variable encoding event types. From a quick look at data, the fatalities and injuries variables would seem logical measures of harm to population health. Potential variables for measuring economic consequences are: PROPDMG, CROPDMG (numbers) and PROPDMGEXP, CROPDMGEXP (factors). Potential location variables include: COUNTY (number), COUNTYNAME (factor), STATE (factor), BGN & END RANGE, AZI, LOCATI (all factors), STATEOFFIC, ZONENAMES (factors), LATITUDE, LONGITUDE, LATITUDE_E, and LONGITUDE_ (all numbers).  

### From the FAQ: 
(https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf)

- Storm data is entered into the database as a distance in miles and a direction on 16-point compass scale from a known location, usually a town or city.   
- Property and crop damage is a best guess made by the National Weather Service using all the data available at the time.   
- For events that include a fatality, there is a code containing the gender, age and fatality location at the end of the event narrative.   
- An Episode is an entire storm system and can contain many different types of events.
An Event is an individual type of storm event. (Thunderstorm Wind, Hail, Tornado and
Flood are events)   

### From STORM DATA PREPARATION INSTRUCTION: 
(https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)  
NATIONAL WEATHER SERVICE INSTRUCTION 10-1605   
AUGUST 17, 2007   
Operations and Services   
Performance, NWSPD 10-16   
STORM DATA PREPARATION  

- Section 2.1.1 contains a Storm Data Event Table which gives the types of events which are recorded. They are associated with three designators: C - County/Parish; Z - Zone; and M - Marine.    
- Section 2.6.1.2 gives a Direct Fatality Location Table.   
- Property damage estimates should be entered as actual dollar amounts.   

Who is the customer? What location data should be used: county, state, zone? The project instructions include: "Consider writing your report as if it were to be read by a government or municipal manager who might be responsible for preparing for severe weather events and will need to prioritize resources for different types of events." I'd imagine that storm episodes do not abide by geopolitical boundaries, so zone might be most approriate for clustering storms. However, since the user would probably be state or county governments, they would want to know if they were either directly effected or in a position to offer neighborly aid.

## Data Processing

### Exploration
Rank EVTTYPE by FATALITIES & INJURIES and then by PROPDMG & CROPDMG. Pie charts? 
How to add location? How many locations are there? Can they be grouped by rankings?

Explore location variables:

```{r location}
summary(df$STATE)
summary(df$COUNTYNAME)
```

Other than 2 "XX" state entries the state entries look fairly complete. County entries have 651640 of 902297 observations (72%) categorized as "other." Looks like state would be a reasonable start.

How many event types occur in each state? Group data by state and count number of event types EVTTYPE for each.

```{r evts_per_state}
require(plyr)
evts_per_state = ddply(df, "STATE", summarise,
               NumEvnts    = length(unique(EVTYPE))
)
max(evts_per_state$NumEvnts)
```

There are a lot of event types for some states (max=143), so might be challenging to visualize a ranking of them all! Let's see how they rank by health harm & economic damage.

```{r sum_by_state_evtype}
require(plyr)
sum_by_state_evtype = ddply(df, c("STATE","EVTYPE"), summarise,
               NumEvtype    = length(EVTYPE),
               EvtypeFatalities = sum(FATALITIES),
               EvtypeInjuries = sum(INJURIES),
               EvtypeTotalHeath = sum(FATALITIES) + sum(INJURIES),
               EvtypePropDmg = sum(PROPDMG), 
               EvtypeCropDmg = sum(CROPDMG),
               EvtypeTotalDmg = sum(PROPDMG) + sum(CROPDMG)
)

health_by_state_evtype = arrange(sum_by_state_evtype[sum_by_state_evtype$EvtypeTotalHeath>0,], STATE, desc(EvtypeTotalHeath))

dmg_by_state_evtype = arrange(sum_by_state_evtype[sum_by_state_evtype$EvtypeTotalDmg>0,], STATE, desc(EvtypeTotalDmg))

```

For each state list top evtype ranked by health or damage.
state Health Damage
name  evtype evtype

## Results
